Mapping Cancer: the Potential of Cartograms and Alternative Map Displays

### 2020-05-17 literature.Rmd
confirmed use and description of statistics in atlas examples

Table 1: <!--\label{tab:measures}--> Common measures for reporting cancer information.

|Measure |Details| Fig. |
|:--------------------|:------------------------------------------------|:--------------------|
|1. Count| Cancer incidence counts| |
|2. Incidence Rate (per 100,000) | The frequency of cases relative to a fixed population size, usually 100,000 people, within a specific period of time XXX [cite it](https://www.cancerresearchuk.org/health-professional/cancer-statistics/cancer-stats-explained/statistics-terminology-explained#heading-Eleven)|  |
|3. IR (Incidence Ratio)| The incidence rate in a region divided by the average  incidence rate for all of the regions| |
|4. Age-Standardized Rate per 100,000| Standardized by age structure or region|   |
|5. Age-Standardized Relative Risk| Standardized by age structure|  |
|6. SIR (Standardized Incidence Ratio)| The observed number of cases in a region divided by the expected number of cases, if that region had the same age-specific rates as a reference population.  |
|7. Excess Hazard Ratio | The estimate of cancer-related mortality within five years of diagnosis. |  |

@@

| Fig. | Atlas | Data | Statistic | Source |
|------|---------------------------------------------------------------------------------------------------------|-----------------------------------------------------|------------------------------------------------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| a | The Environment and Health<br>Atlas of England and Wales<br>[@EnvEnglandWales2] | Sex: Women<br>Year: 2010<br>Cancer: Lung | Relative risk | [Office for National Statistics<br>(ONS) (England) and from the<br>Welsh Cancer Intelligence and<br>Surveillance Unit (WCISU)](http://www.envhealthatlas.co.uk/homepage/) |
| b | Globocan 2018: Estimated Cancer<br>Incidence, Mortality and <br><br>prevalence Worldwide<br>[@Globocan] | Sex: Both<br>Year: 2018<br>Cancer: All invasive | Age standardized<br>incidence rates<br>(per 100,000) | [World Health Organization's<br>International Agency for<br>Research on Cancer](https://www.iarc.fr/<br> |
| c | Atlas of Cancer in Queensland<br> [@QLDcancerAtlas] | Sex: Males<br>Year: 1998 - 2007<br>Cancer: Lung | Relative incidence<br>ratio | [Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/) |
| d | Australian Cancer Atlas<br>[@ACA] | Sex: Males<br>Year: 2010 - 2014<br>Cancer: Lung | Average standardised<br>incidence rate | [Australian Cancer Atlas](https://atlas.cancer.org.au/) |
| e | United States Cancer<br>Statistics: An Interactive<br>Cancer Statistics Website<br> | Sex: Both<br>Year: 2016<br>Cancer: All types | Incidence rate per<br>100,000 | With data from state cancer<br>registries: <br>[Centers for Disease Control and Prevention](https://www.cdc.gov/) |
| f | Map of Cancer Mortality<br>Rates in Spain<br>[@cancerSpain] | Sex: Both<br>Year: 2004 - 2008<br>Cancer: All types | Relative risk | <br>[Map of cancer mortality rates<br>in Spain](https://elpais.com/elpais/2014/10/06/media/1412612722_141933.html |
| g | Atlas of Childhood<br>Cancer in Ontario<br>[@OntarioPediatric] | Sex: Both<br>Year: 1995 - 2004<br>Cancer: All types | Incidence rate per<br>100,00 | <br>[The Pediatric Oncology Group<br>of Ontario Networked Information<br>System](https://www.pogo.ca/research-data/pogonis-childhood-cancer-database/data-requests/) |

@@

Choropleth maps colour each geographic unit to allow map users to measure the value of the statistic [@EI]. Map users contrast the colours in neighbouring areas to understand the spatial distribution. The ColorBrewer system [@CB] and viridis [@viridis] palettes provide effective colour schemes for qualitative, sequential and diverging data. When communicating information using colour, a map creator should use a scheme that has a linear color gradient, with perceptually uniform color spaces that match equal steps in data space with equal steps in the colour space [@PUCS]. 
The use of borders and backgrounds, and their colours, can also change the appearance of the colors representing the value of the statistics [@CB]. These supports can be used to implement a reference point in the colour scheme as well as orient users to the geographic regions.

@@

Maps can also be used to focus on demographic strata, such as age and sex. Some of the digital atlases surveyed allow subsets to be selected for display, The Australian Cancer Atlas [@ACA] allows filtering for the selection of statistics regarding males and females specifically. Similarly, remoteness [@ACA], socioeconomic indicators, such as unemployment rates, poverty rates, and education levels, can be used to filter data, in order to communicate how cancer prevalence varies for different population subgroups. Few atlases provide this level of detail. XXX

Introducing population and demographic information helps to interpret the rates in areas effectively, but there will still be uncertainty around the rates. To address this, a cancer atlas often communicates uncertainty about the value of a statistic. There are several potential sources of uncertainty: sampling error, errors arising from the disease reporting or data collection processes, and uncertainty arising from the statistical modelling or simulation process. The most common measures used to present uncertainty are credible intervals or confidence intervals. Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it is difficult to display effectively. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two common solutions.

>>>>

Table 1: <!--\label{tab:measures}--> Common measures for reporting cancer information. Each measure is given along with a description and the Atlases in Figure 1 that display each measure.

|Measure | Details | Fig. |
|:--------------------|:---------------------------------|:----|
|1. Cancer incidence counts| the number of new cases occurring in a specified period and geographic area [@Globocan]| |
|2. Incidence Rate (per 100,000) | The frequency of cases relative to a fixed population size, usually 100,000 people, within a specific period of time [@Globocan] | e, g |
|3. Incidence Ratio | The incidence rate in a region divided by the average incidence rate for all of the regions| c |
|4. Age-Standardized Incidence Rate (per 100,000) | The weighted mean of the age-specific incidence rates; the weights are taken from population distribution of the standard population. The most frequently used standard population is the World Standard Population. [@Globocan] | b, d |
|5. Age-Standardized Relative Risk | The relative risks represent the risk of an area (ward) relative to average risk, adjusted for age, deprivation and chance fluctuations due to small numbers. [@EnvEnglandWales2] | a, f |

Table describes the common measures used to present cancer incidence in atlases. The calculations used to produce these measures often are also applied to mortality statistics.The Excess Hazard Ratio and Standardized Incidence Ratio are additional measures that can be found in atlases, but are not used in the images displayed in Figure 1.

@@

| Fig. | Atlas | Data | Statistic | Source |
|------|------------|-----------|----------------------|-----------------------------|
| a | The Environment and Health<br> Atlas of England and Wales<br> [@EnvEnglandWales2] | Sex: Women<br> Year: 2010<br> Cancer: Lung | Relative risk | [Office for National Statistics<br> (ONS) (England) and from the<br> Welsh Cancer Intelligence and<br> Surveillance Unit (WCISU)](http://www.envhealthatlas.co.uk/homepage/) |
| b | Globocan 2018: Estimated Cancer<br> Incidence, Mortality and <br> <br> prevalence Worldwide<br> [@Globocan] | Sex: Both<br> Year: 2018<br> Cancer: All invasive | Age standardized<br> incidence rates<br> (per 100,000) | [World Health Organization's<br> International Agency for<br> Research on Cancer](https://www.iarc.fr/<br>  |
| c | Atlas of Cancer in Queensland<br>  [@QLDcancerAtlas] | Sex: Males<br> Year: 1998 - 2007<br> Cancer: Lung | Relative incidence<br> ratio | [Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/) |
| d | Australian Cancer Atlas<br> [@ACA] | Sex: Males<br> Year: 2010 - 2014<br> Cancer: Lung | Average standardised<br> incidence rate | [Australian Cancer Atlas](https://atlas.cancer.org.au/) |
| e | United States Cancer<br> Statistics: An Interactive<br> Cancer Statistics Website<br>  | Sex: Both<br> Year: 2016<br> Cancer: All types | Incidence rate per<br> 100,000 | With data from state cancer<br> registries: <br> [Centers for Disease Control and Prevention](https://www.cdc.gov/) |
| f | Map of Cancer Mortality<br> Rates in Spain<br> [@cancerSpain] | Sex: Both<br> Year: 2004 - 2008<br> Cancer: All types | Relative risk | <br> [Map of cancer mortality rates<br> in Spain](https://elpais.com/elpais/2014/10/06/media/1412612722_141933.html |
| g | Atlas of Childhood<br> Cancer in Ontario<br> [@OntarioPediatric] | Sex: Both<br> Year: 1995 - 2004<br> Cancer: All types | Incidence rate per<br> 100,00 | <br> [The Pediatric Oncology Group<br> of Ontario Networked Information<br> System](https://www.pogo.ca/research-data/pogonis-childhood-cancer-database/data-requests/) |

@@

Choropleth maps colour each geographic unit to allow map users to measure the value of the statistic [@EI]. Map users contrast the colours in neighbouring areas to understand the spatial distribution. The ColorBrewer system [@CB] and viridis [@viridis] palettes provide effective colour schemes for qualitative, sequential and diverging data. When communicating information using colour, a map creator should use a scheme that has a linear color gradient, with perceptually uniform color spaces that match equal steps in data space with equal steps in the colour space [@PUCS]. It is possible to allow for data to progress uniformly in both positive and negative directions from a mid point, such as the mean of the data. These diverging colour schemes pair two sequential schemes that use a common light colour at the mid point, each sequential scheme progresses to a dark hue at the extreme value in each direction [@DCS].
The use of borders and backgrounds, and their colours, can also change the appearance of the colors representing the value of the statistics [@CB]. These supports can be used to implement a reference point in the colour scheme as well as orient users to the geographic regions.

@@

Maps can also be used to focus on demographic strata, such as age and sex. Some of the digital atlases surveyed allow subsets to be selected for display, for example the Australian Cancer Atlas [@ACA] allows filtering for the selection of statistics regarding males and females specifically.

Introducing population and demographic information helps to interpret the rates in areas effectively, but there will still be uncertainty around the rates. To address this, a cancer atlas often communicates uncertainty about the value of a statistic. There are several potential sources of uncertainty: sampling error, errors arising from the disease reporting or data collection processes, and uncertainty arising from the statistical modelling or simulation process. The most common measures used to present uncertainty are credible intervals or confidence intervals. Displaying the uncertainty associated with reported statistics is a vital feature of a cancer map, but it is difficult to display effectively, the Australian Cancer Atlas [@ACA] uses transparency to communicate uncertainty. Providing an adjacent map or overlaying maps with symbols [@VSSDCUC] are two common solutions.
